---
title: "README"
output: html_document
---

## Location of the run_analysis.R script
The Run_analysis script goes in the UCI directory which also contains the train and test folders

```{r, message=FALSE, error=FALSE, warning=FALSE, results="hide"}
setwd("E:\\Dev\\R\\Coursera\\3 Cleaning\\Wearable-Computing\\")
```


## Setting up packages and environment
```{r, message=FALSE, error=FALSE, warning=FALSE, results="hide"}
# Load packages we are going to work with
library(lubridate)      # Working with dates
library(dplyr)          # Working with tables
library(ggplot2)        # Working with plots
library(tidyr)          # Tidying data
library(plyr)
library(data.table)

# Set the locale the english
Sys.setlocale("LC_TIME", "English")
```



## 1. Merging the training and the test sets to create one data set.

### Reading the activity rows
```{r}

# Read the activity rows for test
activitytest<-read.table("./test/y_test.txt", header=F)

# Read the activity rows for train
activitytrain<-read.table("./train/y_train.txt", header=F)

# Add the test and train data together
activity<-rbind(activitytest, activitytrain)

# Check the rows structure
dim(activity) #10299 activity rows

# Create a column name
colnames(activity)<- c("activityid")

```


## Reading the subject rows
```{r}
# Read the subject rows for test
subjecttest<-read.table("./test/subject_test.txt", header=F)

# Read the subject rows for train
subjecttrain<-read.table("./train/subject_train.txt", header=F)

# Add the test and train data together
subject<-rbind(subjecttest, subjecttrain)

# Check the rows structure
dim(subject)

# Create a column name
colnames(subject)<- c("subject")
```

### Reading the featurelabels
```{r}
# Read the featurelabels for the columns of the test and train data
featurelabels<-read.table("features.txt", header=F, sep=" ")

# Check the structure
dim(featurelabels) #561 column labels

# Create column names
colnames(featurelabels)<-c("featurenr", "featurename")
```

### Read the test and train data
```{r}
# Read the x-test feature data
featurestest<-read.table("./test/x_test.txt", header=F)

# Read the x-train feature data
featurestrain<-read.table("./train/x_train.txt", header=F)

# Add the test and train data together
features<-rbind(featurestest, featurestrain)

# Check the structure
dim(features) #561 columns
```

### Create the combined dataset
```{r}
# Create column names for the featurelabels
# Ensure that the column names are unique, otherwise we'll encounter errors later when merging
colnames(features) <- make.names(featurelabels$featurename, unique=TRUE)

# add the activity, subject feature data together
data<-cbind(activity, subject, features)

# Recheck the structure
dim(data)

```




# 2. Extract only the measurements on the mean and standard deviation for each measurement. 
```{r}

# Create a new table containing only columns with subjects, activityid, mean and std
datameanstd <- select(data,matches("subject|activityid|mean|std"))

# Check the structure
dim(datameanstd)

```


# 3. Uses descriptive activity names to name the activities in the data set
```{r}

# Read the labels for the activities
activitylabels<-read.table("activity_labels.txt", header=F, sep=" ")

# Check the structure
head(activitylabels)

# Create column names for activity labels
colnames(activitylabels)<- c("activityid","activity")

# Add the activity label to the dataset using a merge on activityid
data <- merge(x=datameanstd, y=activitylabels, by="activityid")

# Check that activity has been merged correctly
unique(data[,c("activity")])

# Exclude the activityid field
data <- select(data, -activityid)

# Reorder the columns so that the dataset starts with subject and activity
data<-select(data, subject, activity, 2:87)

# Check the structure
dim(data)
str(data)

```



# 4. Appropriately labels the data set with descriptive variable names. 
```{r}

# Get the column names and make them unique
colnames <-colnames(data)
colnames <- make.names(colnames, unique=TRUE)

#Cleanup the variable names by replacing characters
colnamesclean<-gsub("-", " ", colnames) #Rplace - with a space
colnamesclean<-gsub("\\.", " ", colnamesclean) #Replace . with a space
colnamesclean<-gsub("\\  ", " ", colnamesclean) #Replace . with a space
colnamesclean<-gsub("\\  ", " ", colnamesclean) #Replace . with a space
colnamesclean<-gsub("\\  ", " ", colnamesclean) #Replace . with a space
colnamesclean<-gsub("tBody", "Body", colnamesclean) #Remove the t
colnamesclean<-gsub("tGravity", "Gravity", colnamesclean) #Remove the t
colnamesclean<-gsub("fBody", "Body", colnamesclean) #Remove the f
colnamesclean<-gsub("BodyBody", "Body", colnamesclean) #Remove double Body
colnamesclean<-gsub("^\\s+|\\s+$", "", colnamesclean) #Strip leading and trailing spaces

# Recreate the column names for the dataset
colnames(data) <- colnamesclean

# Check the structure
str(data)

```



## 5. Create tidy dataset from step 4
From the data set in step 4, create a second, independent tidy data set with the average of each variable for each activity and each subject
```{r}

# Create a datafram table (Dplyr)
tidy <- tbl_df(data)

# Create unique column names, otherwise the summary will give errors
colnames(tidy) <- make.names(colnames(tidy) , unique=TRUE)

# Group the data by subject and activity
tidygroup <-group_by(tidy, subject, activity)

# Calculate the mean for all features using a Dplyr function
tidymean <- summarise_each(tidygroup, funs(mean))

# Reapply the clean column names
colnames(tidymean) <- colnamesclean

# Check the first 10 rows and 6 columns
tidymean[1:10, 1:6]

# Create tidy dataset from step 5
write.table(tidymean, file="tidy.txt", row.names=FALSE, col.names=TRUE, sep="\t", quote=TRUE)


```
